\textbf{Analysis of Greenland narwhals's behavioural responses to ship noise and airguns exposure using varying coefficients correlated velocity models}

### 2024-04-15 Latex/main.tex
Commit Latex and simulation study changes

%%PACKAGES
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage[affil-it]{authblk}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

@@

%context and literature review, objective of the paper, new contributions
% ARCTIC , narwhals etc
Assessment of human disturbance on animal behaviour has become a major concern in the ecology community. Both instantaneous and long-term changes in animal behaviour have been studied by taking advantage of the large amount of wildlife data now available. One of the commonly accepted method to analyse such behavioural changes due to human activities is to perform "controlled exposure experiments". Animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, sound exposure level, temperature, pressure, heart beat, stroke rate...  
Several studies involving beaked whales \parencite{tyack_beaked_2011}, common sole larvaes \parencite{bolle_common_2012}, narwhals \parencite{heide-jorgensen_behavioral_2021} and many other species have already been done. Marine mammals are particularly challenging to track since GPS data can only be retrieved when they emerge at the surface of the sea and a satellite is near enough. Moreover, some species can dive at great depth, so that the equipment must keep working in such extreme environment \parencite{cioffi_trade-offs_nodate}. In this paper, a data-set of east Greenland narwhals's movement under controlled exposure experiments will be analysed. \\

As a consequence of climate change and decrease in sea ice coverage, the Greenland ecosystem is likely to be more and more exposed to anthropogenic activity in the next decades \cite{heide-jorgensen_behavioral_2021}. Shipping activities as well as sonar and seismic surveys are a few instances of new disturbances that could alter animals's habits.  Hence, there is an urge for more knowledge about the impact this could have on endemic species such as the narwhals. This marine mammal is one of the most sensitive to sound exposure as it uses echolocation to orient itself beneath the water \cite{tervo_narwhals_2021}. Moreover, recent studies showed that the population of narwahls in Southeast Greenland is already in decline \cite{garde_biological_2022}. The effects of sound and ship exposure on its diving and foraging behaviour have already been the subject of several studies \cite{heide-jorgensen_behavioral_2021,ngo_understanding_2019}. However, their horizontal movement has for now been analyzed with discrete time methods such as Markov chains  \cite{heide-jorgensen_behavioral_2021}. A covariate representing the level of sound received by the narwhals was used to model the transition probabilities of the Markov Chain, and it was concluded that narwhals effectively tried to avoid the ship by moving fast toward the shore.\textcolor{red}{add more detailed explanations}. Here, we use a continuous time model to analyse the GPS positions and try to corroborate this analysis. The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\

We propose to use a stochastic differential equation with varying coefficient to model the GPS tracks of the individuals. The continuous time correlated random walk (also known as integrated Ornstein-Uhlenbeck, or correlated velocity model) that has been introduced in \cite{johnson_continuous_2008} is one of the most common model one can find in the literature to analyse such data. It has been  widely used in ecology to model 2D animal movement \cite{gurarie_correlated_2017,albertsen_generalizing_2019} \textcolor{red}{add more references}, and even 3D movement \cite{gurarie_estimating_2011}.

 Following \cite{gurarie_correlated_2017}, this model can be written
\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-\frac{1}{\tau} (v(t) -\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: CVM equation}
\end{equation}
$v(t)$ is the horizontal 2D velocity at time $t$ and $x(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, and $\mu$ is the long term velocity. This equation is a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. \\

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process. However, this phenomenon is not taken into account in model~\ref{eq: CVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recent papers including constraints are \cite{hanks_reflected_2017} and \cite{russell_spatially_2018}.
 In \cite{hanks_reflected_2017}, a reflected version of the CVM is considered to study sea lion telemetry data.
For the simulation, points ending up in land are projected on the coastline.  Estimation of the SDE parameters relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. This requirement is met by linearly interpolating between the discrete time observations of the process.
In \cite{russell_spatially_2018}, a repulsive potential function is used to constrain ants movement within a box.
Usually, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For the narwhals, it was concluded from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.

To add a rotational component, we propose to use the rotational version of the CVM \cite{gurarie_correlated_2017,albertsen_generalizing_2019}.
\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-A(v(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega \\
	\omega & \frac{1}{\tau}
\end{pmatrix}$.
The new parameter $\omega$ is an angular velocity that controls how fast the velocity vector should rotate. Notice that if $\omega=0$ this is simply equation \ref{eq: CVM equation}.

The novelty of this paper consists in estimating the parameters $\tau$ and $\omega$ as smooth functions of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This is a way to make the drift in the velocity equation dependent on the location process $x$. \\
Such a model can be estimated in the framework of the \texttt{R} package \texttt{smoothSDE}. To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuous_2008}. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.

On top of this, the parameter $\nu$ will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
The effect of sound exposure is assessed both in the constrained and unconstrained frameworks.

To summarize, the main contributions of this paper are
\begin{itemize}
	\item Derivation of a state-space model for the RACVM defined in \cite{gurarie_correlated_2017}, and addition of this method in the framework of \texttt{smoothSDE}, to enable the use of smooth parameters $\nu$,$\tau$,$\omega$ and $\mu$, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a Constrained Rotational Correlated Velocity Model (CRCVM) for simulation study. Deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the Distance to the ship.
\end{itemize}

Note that simpler SDE models as Brownian motion or Ornstein-Uhlenbeck with varying coefficient already proved to be convenient at assessing behavioural responses by examining deviation from a baseline model that represents the animal behaviour under normal conditions. For instance, it was used in \cite{michelot_continuous-time_2022} to estimate avoidance behaviours of beaked whales after exposure to military sonar systems.

Eventually, while a lot of results are known about the existence and properties of constrained diffusions, their application in the context of animal motion remains relatively underdeveloped \cite{brillinger_simulating_2003}. This text aims to contribute to bridging this gap.
\newpage

\section{Movement data of Greenland endemic narwhals}
%describe how the data was obtained, the geography, type of whales, year, exposure to boats etc..
%trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
The dataset analysed in this paper has already been the subject of several studies, focusing mainly on the diving behaviour of the narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021,ngo_understanding_2019}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters. The Scoresby Sound fjord system
is known to be the summer residence for an isolated population of narwhals.
Here we recall briefly how the data has been retrieved. For more details about the study area and the tagging of the animals, the reader may refer to \cite{heide-jorgensen_behavioral_2021}. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every 80 s,
and the GPS navigation system recorded the
location of every shot. 
Data includes narwhals's latitude and longitude, distance of the narwhals relative to the ship in metres, GPS time, distance to the shore in metres. GPS positions are known only at specific times when the narwhals get close to the sea surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.\textcolor{red}{add histograms of time steps}\\

@@

The land geometry for this specific region in South-East Greenland was gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
For each GPS measurement time $t_j$, the nearest point $p(t_j)$ on the shoreline was computed and the distance to the shore was determined as the distance to this point. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This was primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points were ignored. Then, the exposure to shore covariate was defined with the following formula
\begin{equation}
\left\{
\begin{array}{ll}
	 E_{shore}=0 & \mbox{if } D_{shore}>D_0 \\
	E_{shore}=\frac{1}{D_{shore}} & \mbox{if } D_{shore} < D_0
\end{array}
\right.
\label{eq: exp shore definition}
\end{equation}
where the threshold $D_0$ was chosen equal to $0.5$ km.
The angle between the narwhal's heading and the shoreline was computed as the angle $\theta$ between each step vector $x(t_j+\Delta_j)-x(t_j)$ and the vector $n(t_j)=x(t_j)-p(t_j)$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,\frac{-\pi}{2}\right[$ indicates movement toward the shore. 
The distance from the ship was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}. The covariates $E_{shore}$ and $\Theta$ will only be used in the constrained SDE models.

@@

		%nodes
		\node[left] at (0,0) {$\partial \mathcal{D}$};
		\node[left] at (0,2) {$p(t_j)$};
		\node[left] at (1.2,2.8) {$\vec{v}(t_j)$};
		\node[below,red] at (3,2) {$\vec{n}(t_j)$};

		\node[above,orange] at (2.3,2.3) {$\theta(t_j)$};
	\end{tikzpicture}
	\caption{Example of nearest shore point and angle $\theta$}
	\label{fig: illustration nearest shore points}
\end{figure}
\begin{figure}[H]
    \centering
    \begin{subfigure}{0.48\textwidth}
    \centering
     \includegraphics[scale=0.45]{images/data_exploration/Dship_histo.png}
    \caption{Histograms of the distance to the ship \\ for each narwhal}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[scale=0.45]{images/data_exploration/Dshore_histo.png}
    \caption{Histograms of the distance to the shore \\ for each narwhal}
    \end{subfigure}
    \caption{Distance to ship and distance to shore covariates}
    \label{fig: distance to shore and distance to ship histograms}
\end{figure}

The exposure to ship covariate was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D_{ship}$ is not known when the narwhal is not in line of sight with the boat, so that exposures are set to $0$ in this case. The closer the narwhal is to the ship, the greater the exposure. This is a proxy for sound exposure levels received by the narwhals. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.All the covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.

@@

\subsection{Rotational correlated velocity model (RCVM)}
\label{section: RCVM}

The more general equation for a rotational correlated velocity model is \cite{albertsen_generalizing_2019}:

\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-A(v(t)-\mu)dt+\Sigma dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$.
Here, $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity. This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2018}.
There is an explicit form for the gaussian transition density of the velocity process \cite{albertsen_generalizing_2019}, where the covariance matrix is expressed using a Kronecker sum.

It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$.
This diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$ to give biological meaning to the parameters \cite{gurarie_correlated_2017}.These simplifications produce the equation~\ref{eq: RACVM equation}.

When parameters $\tau$, $\mu$ and $\nu$ are constant over time, both the velocity and the position processes can be solved explicitly.
The transition density for the velocity process is such that 
\begin{equation}v(t+\Delta) \vert v(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

@@

\textcolor{red}{Talk about the consistency and asymptotic properties of such estimators}

@@

\section{Precisions about the statistical models}

In the baseline mixed effect model, the explicit model matrices are

@@

\section{More simulation study}

>>>>

%%PACKAGES
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage[affil-it]{authblk}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

%CITATIONS APPEARANCE
\DeclareCiteCommand{\cite}[\mkbibbold]
{\usebibmacro{prenote}}
{\usebibmacro{citeindex}%
	\printtext[bibhyperref]{\usebibmacro{cite}}}
{\multicitedelim}
{\usebibmacro{postnote}}

\DeclareFieldFormat{boldcite}{\mkbibbold{#1}}

\DeclareCiteCommand{\cite}[\mkbibbold]
{\usebibmacro{prenote}}
{\printtext[brackets]{\usebibmacro{cite}}}
{\multicitedelim}
{\usebibmacro{postnote}}

@@

% Underwater noise in Arctic : lower than in other environments, anthropogenic nois : vessel traffic, ice breakers, drilling, sonars, seismic airguns for oil and gas exploration. (Halliday 2020)
% vessel traffic likely to increase due to the melting of the ice cap
% Largest impact on underwater noise due to seismic airguns (Halliday 2020)
%Marine mammals sound exposure effects : behavioral and physical impacts (stranding events), more difficult to assess behavioral effect (southall 2007, 2021)
%Sound threshold, regulation, research recommendations : behavioral response
%Behavioral response study, control exposure experiments (examples: pile driving, seismic airguns, sonar)

% Studies about a lot of species : narwhals, toothed whale,  mid frequency mammals
% Greenland narwhals studies
% effects on dive, vocalization and horizontal movement
% data, diffusion models 
% spatial constraints
%context and literature review, objective of the paper, new contributions
 %% Southeast Greenland narwhals population in decline

Anthropogenic underwater noise in the Arctic environment causes disturbances in the endemic wildlife. More precisely, marine mammals, whose vital functions highly depend on sound perception for communication and orientation, are likely to react to such disturbances. The Arctic Council report \cite{halliday_underwater_2020} emphasized the need for more knowledge about the impact of underwater noise on Arctic marine mammals. The main sources of noise in the region are vessel traffic, ice breakers, drilling, sonars and seismic airguns used for oil and gas exploration. Among those sources, an increase in vessel traffic is anticipated due to the melting of the ice cap, and seismic airguns are reported to have the largest impact on underwater noise. Two types of consequences can be considered : injury and behavior disturbance. While physical harm is straightforward to assess and mitigate by determining species-specific received sound level thresholds, no canonical method is expected to be found to examine behavior disturbances, due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions.\\
Research recommendations include a particular focus on measurement of behavioral reactions to various type of sound exposure \cite{southall_marine_2008,southall_marine_2019}. Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations... Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises \cite{tyack_beaked_2011,cioffi_trade-offs_2022}, belugas \cite{martin_exposure_2023}, sperm whales \cite{madsen_quantitative_2006}, blue whales \cite{friedlaender_preymediated_2016}, narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021} and many other marine mammals. \\
A severity scale was defined in \cite{southall_marine_2008} to help assessing which behavioral responses are more likely to alter a population's capacity to survive, reproduce or forage. Changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations  were ranked between 4 and 7 on a scale out of 9 (9 being stranding events that can directly lead to the death of the animal) depending on the magnitude of the changes and their duration.\\

So far, the studies about narwhals behavioral responses to sound exposure proved that they exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}. Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source \cite{tervo_narwhals_2021}. These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}. Resulting GPS positions of the narwhals have been analyzed with discrete time methods, typically relying on Markov chains. Since GPS positions are retrieved irregularly in time, each narwhal track was linearly interpolated to obtain a position every second, and the positions were then classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second. An exposure covariate defined as the inverse of the distance to the sound source was used to model the transition probabilities of the Markov Chain, and it was found that the probabilities of being in states "Close to the shore" and "Moving toward the shore" increased with exposure to the sound, while probability of being in state "Far from the shore" decreased. In this paper, we adopt a different approach to describe horizontal movement and make use of a continuous time model to analyse the GPS positions. The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\

A stochastic differential equation with time-varying coefficient is employed to describe the GPS tracks of the individuals \cite{michelot_varying-coefficient_2021}. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It has been used extensively to analyze animal movement in two or even three dimensions \cite{johnson_continuoustime_2008,gurarie_correlated_2017,alt_correlation_1990,gurarie_estimating_2011,albertsen_generalizing_2018}, and can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
$V(t)$ is the horizontal 2D velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates. Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.\\

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process. However, this phenomenon is not taken into account in model~\ref{eq: RACVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recent research has been conducted to include constraints in these type of models.
A reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
Simulation of constrained paths was achieved by projecting on the coastline the  points that eventually reached land. Estimation of the SDE parameters relied on computationally intensive MCMC algorithm (Metroplis-Hastings), and required high frequency data. This requirement was met by linearly interpolating between the discrete time observations of the process. Another idea is to add a repulsive potential function in the drift. This has been done to constrain ants movement within a box \cite{russell_spatially_2018}, but requires very simple boundaries to get an explicit expression of the potential function.\\
Moreover, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For narwhals, it was observed from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline. The novelty of this paper consists in estimating the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector, therefore making the drift in the velocity equation dependent on the location process $x$ and the boundary of the domain. Note that in this framework, the special case $\omega=0$ would be the unconstrained correlated velocity model of \cite{johnson_continuoustime_2008}.  \\ %More biblio about estimation methods
The angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE}. To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\

Sound exposure will be considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}. This effect will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship. This choice of covariate is motivated by previous work in \cite{heide-jorgensen_behavioral_2021}. A response model will be compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
The effect of sound exposure is assessed both in the constrained and unconstrained ($\omega=0$) frameworks.

To summarize, the main contributions of this paper are
\begin{itemize}
	\item Derivation of a state-space model for the so called "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}) \cite{gurarie_correlated_2017}, and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a Constrained Rotational Correlated Velocity Model (CRCVM) for simulation study. Deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

Section 2 gives an overview of the narwhal data available for the analysis. Then in section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models. Section 4 is dedicated to a comprehensive simulation study, whereas in the last section, the results of the application to the narwhal data are examined.

\section{Movement data of Greenland endemic narwhals}
%describe how the data was obtained, the geography, type of whales, year, exposure to boats etc..
%trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
As already mentioned, the dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters. The Scoresby Sound fjord system
is known to be the summer residence for an isolated population of narwhals.
Here we recall briefly how the data has been retrieved. For more details about the study area and the tagging of the animals, the reader may refer to \cite{heide-jorgensen_behavioral_2021}. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every 80 s,
and the GPS navigation system recorded the
location of every shot. 
Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres. GPS positions are known only at specific times when the narwhals get close to the sea surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours. Figure \ref{fig:alltimestepshisto} shows the distribution of time steps in the data.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/data_exploration/all_time_steps_histo}
	\caption{Histogram of time step between consecutive GPS positions}
	\label{fig:alltimestepshisto}
\end{figure}

@@

The land geometry for this specific region in South-East Greenland was gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
Supposing there are $n$ observations, for $j \in \{1,\cdots,n\}$, the GPS position obtained at time $t_j$ is denoted $y(t_j)$, and the nearest point on the shoreline is denoted $p(t_j)$. The distance to the shore, denoted $D_{shore}(t_j)$ in the sequel, is determined as the distance to this point. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This was primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points were ignored. Then, the exposure to shore covariate was defined with the following formula
\begin{equation}
\left\{
\begin{array}{ll}
	 E_{shore}(t_j)=0 & \mbox{if } D_{shore}(t_j)>D_0 \\
	E_{shore}(t_j)=\frac{1}{D_{shore}(t_j)} & \mbox{if } D_{shore}(t_j) \leq D_0
\end{array}
\right.
\label{eq: exp shore definition}
\end{equation}
where the threshold $D_0$ was chosen equal to $0.5$ km.
The angle between the narwhal's heading and the shoreline is estimated as the angle $\Theta(t_j)$ between the observed empirical velocity $\hat{v}(t_j)=\frac{y(t_j+\Delta_j)-y(t_j)}{\Delta_j}$, where $\Delta_j=t_{j+1}-t_j$, and the vector $n(t_j)=y(t_j)-p(t_j)$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta(t_j)=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta(t_j) \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore. 
The distance from the ship, denoted $D_{ship}(t_j)$, was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}. The covariates $E_{shore}$ and $\Theta$ will only be used for the smooth angular velocity parameter $\omega$ in the constrained SDE models.

@@

		%nodes
		\node[left] at (0,0) {$\partial \mathcal{D}$};
		\node[left] at (0,2) {$p(t_j)$};
		\node[left] at (1.2,2.8) {$\hat{v}(t_j)$};
		\node[below] at (1.8,2) {$y(t_j)$};
		\node[below,red] at (3,2) {$\vec{n}(t_j)$};

		\node[above,orange] at (2.3,2.3) {$\Theta(t_j)$};
	\end{tikzpicture}
	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain.}
	\label{fig: illustration nearest shore points}
\end{figure}
\begin{figure}[H]
    \centering
    \begin{subfigure}{0.48\textwidth}
    \centering
     \includegraphics[scale=0.45]{images/data_exploration/Dship_histo.png}
    \caption{Histograms of the distance to the ship \\ for each narwhal}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[scale=0.45]{images/data_exploration/Dshore_histo.png}
    \caption{Histograms of the distance to the shore \\ for each narwhal}
    \end{subfigure}
    \caption{Distance to ship and distance to shore covariates}
    \label{fig: distance to shore and distance to ship histograms}
\end{figure}

The exposure to ship covariate, denoted $E_{ship}$, was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D_{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case. This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source. This covariate is meant to be a proxy for sound exposure levels received by the narwhals. Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}. All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.

@@

\subsection{Standard Rotational correlated velocity model (RCVM)}
\label{section: RCVM}

A general formulation for a rotational correlated velocity model is \cite{albertsen_generalizing_2018}:

\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(V(t)-\mu)dt+\Sigma dW(t)

	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity. This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2019}.
There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.

It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$ \cite{johnson_continuoustime_2008}.
The diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$, where $\nu$ is homogeneous to a velocity. This allows easier biological interpretation  of the parameters in the equation \cite{gurarie_correlated_2017}. These simplifications produce the equation~\ref{eq: RACVM equation}.

When parameters $\tau$, $\mu$ and $\nu$ are constant over time, both the velocity and the position processes can be solved explicitly.
The transition density for the velocity process is such that 
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

@@

%Talk about the consistency and asymptotic properties of such estimators

@@

\section{Precisions about the statistical models}
\label{section: precisions about statistical models}
In the baseline mixed effect model, the explicit model matrices are

@@
